Title: Predicting Human Activity Using Smartphone Movement Measurements using a Random Forest model
========================================================
```{r loadData, cache=TRUE, eval=TRUE, echo=FALSE, include=FALSE}
proj_dir = "~/classes/dataAnalysis002/proj2"
setwd(proj_dir)

# remote data
rdaUrl <- "https://spark-public.s3.amazonaws.com/dataanalysis/samsungData.rda"
descUrl <- "http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones"

dataDir <- 'data'
if (!file.exists(dataDir)) {
  dir.create(dataDir)
}

getDownloadedFile <- function(url, dir=dataDir) {
  file <- paste(dir, basename(url), sep='/')
  if (!file.exists(file)) {
    download.file(url, file, method='curl')
  }
  invisible(file)
}

rdaFile <- getDownloadedFile(rdaUrl)
load(rdaFile, verbose=TRUE)

rdaDownloadTime <- file.info(rdaFile)$ctime

#######
# munge data
#######
# clean up column names making them syntactic names for ease of use
v <- gsub("\\()",     "",   names(samsungData)) # remove () in names
v <- gsub("-",        ".",  v)                  # remove dashes
v <- gsub("[\\(\\)]", "..", v)                  # remove parantheses
v <- gsub(",",        "_",  v)                  # remove commas
names(samsungData) = v

# clean up data by making subject and activity factors
samsungData$subject = factor(samsungData$subject)
samsungData$activity = factor(samsungData$activity)

subject_ndx <- which(names(samsungData)=='subject')
activity_ndx <- which(names(samsungData)=='activity')
factor_ndx <- c(subject_ndx, activity_ndx)

# create user-scaled data
l <- split(samsungData[,-factor_ndx], samsungData$subject)
l <- lapply(l, function(df) { scale(df) })
ssData <- data.frame(do.call('rbind',l))
names(ssData) = paste0("ss.", names(ssData))
allData = cbind(samsungData, ssData)


```

Introduction:
-------------

As of 2013, 56% of American adults now own a smartphone[^1]. Most of these smartphones
have embedded accelerometers and gyroscopes which detect movement of the phone.
In many domain spaces (e.g., exercise detection), it would be nice to determine 
what a phone user was doing based on these measurements captured passively by the phone.

Based on an experiment where volunteers performed a variety of daily activities
with a phone on their waist, I have developed a Random Forest prediction model
to determine what a user's activity is based on a set of numeric movement 
measurements captured by a phone. This model trained on the results of 20
volunteersTesting showed this model to be .

We have developed a Random Forest prediction model  on a data set from a 30
volunteer study coverin6 daily activities

Methods:
-------

_Data Collection_

Smartlab-Non Linear Complex Systems Labroratory at University of Genova, Italy, 
ran an experiment with 30 volunteers performing six daily activities {WALKING, 
WALKING UPSTAIRS, WALKING DOWNSTAIRS, SITTING, STANDING, LAYING} while wearing a
smartphone, Samsung Galaxy S II, on the waist.  During these activities, the phone's
embedded accelerometer and gyroscope captured 3-axial linear acceleration and 3-axial
angular velocity measurements.  These measurements were processed to create a data set
comprising of 561 separate numeric attributes (e.g., mean, standard deviation, 
minimum, and maximum body acceleration in the X, Y, and Z direction) along 
with the subject and activity identifiers. This data set and the original experiment is described 
on the UCI Machine Learning Repository [^2], where the original data set can be found [^3].

```{r eval=TRUE, echo=FALSE, include=FALSE}
tbl_subj = table(samsungData$subject)
tbl_act = table(samsungData$activity)
tbl_subj_act = table(samsungData$subject, samsungData$activity)

activities = toupper(gsub("(up|down)$", " \\1stairs", 
                          gsub("^walk", "walking",
                               levels(samsungData$activity))))
```
For this analysis, the data set was downloaded from a link on Coursera Data 
Analysis Project 2 web page [^4] on `r strftime(rdaDownloadTime, '%F %T %Z')` using 
the R programming language. This project data set contains approximately 70% of the 
original instances, consisting of `r nrow(samsungData)` instances from 
`r nlevels(samsungData$subject)` subjects.  Each subject was measured between 
`r paste0(collapse=" and ", range(tbl_subj))` times; each activity was performed between
`r paste0(collapse=" and ", range(tbl_act), rep(" (", 2), activities[c(which.min(tbl_act), which.max(tbl_act))], rep(")", 2))`  
times; and each subject was measured performing each activity between 
`r paste0(collapse=" and ", range(tbl_subj_act))` times.

All `r ncol(samsungData)-2` numeric attributes are present for each instance in
the data set as well as the 2 categorical attributes: subject 
and activity.

`r nlevels(samsungData$subject)`
`r paste(sort(gsub("(up|down)$", " \\1stairs", gsub("^walk", "walking", levels(samsungData$activity)))), collapse=", ")`

This data set consists of `r nrow(samsungData)` events with `r ncol(samsungData)-2` numeric attributes and 2 categorical attributes: subject and activity.
There are approximately `r round(mean(table(samsungData$subject)))` events per subject and
`r round(mean(table(samsungData$activity)))` events per activity.

_Exploratory Analysis_


After examining the data provided with several techniques such as Singular Value Decomposition



[^1]: http://www.pewinternet.org/~/media/Files/Reports/2013/PIP_Smartphone_adoption_2013.pdf

[^2]: `r descUrl`

[^3]: http://archive.ics.uci.edu/ml/machine-learning-databases/00240/

[^4]: `r rdaUrl`, downloaded on `r strftime(rdaDownloadTime, '%F %T %Z')`

