Title: In addition to FICO Score, increased loan length and loan amount are associated with increased loan rates
========================================================

```{r loadData, cache=TRUE, eval=TRUE, echo=FALSE, include=FALSE}
setwd("~/classes/dataAnalysis002/proj1")

# remote data
dataCsvUrl <- "https://spark-public.s3.amazonaws.com/dataanalysis/loansData.csv"
dataRdaUrl <- "https://spark-public.s3.amazonaws.com/dataanalysis/loansData.rda"
codebookUrl <- "https://spark-public.s3.amazonaws.com/dataanalysis/loansCodebook.pdf"
dataDir <- 'data'
if (!file.exists(dataDir)) {
  dir.create(dataDir)
}

getDownloadedFile <- function(url, dir=dataDir) {
  file <- paste(dir, basename(url), sep='/')
  if (!file.exists(file)) {
    download.file(url, file, method='curl')
  }
  invisible(file)
}

getDownloadedFile(dataCsvUrl)
getDownloadedFile(codebookUrl)

dataRdaFile <- getDownloadedFile(dataRdaUrl)
load(dataRdaFile, verbose=TRUE)
# loads loansData

##########
# munge data
##########
# store off unmunged data
rawData = loansData

loansData$Interest.Rate = as.numeric(sub('%$','', as.character(loansData$Interest.Rate)))
loansData$Debt.To.Income.Ratio = as.numeric(sub('%$','', as.character(loansData$Debt.To.Income.Ratio)))

ficoRangeMeans <- sapply(lapply(strsplit(levels(loansData$FICO.Range), '-'), as.numeric), mean)
loansData$FICO <- ficoRangeMeans[unclass(loansData$FICO.Range)]
# loansData$Loan.Length <- as.integer(sub(" months$", "", unclass(loansData$Loan.Length)))

empLengthLevels = levels(loansData$Employment.Length)
empLengthNumeric = as.numeric(gsub('n/a', '0',
                                   gsub('^< 1$','0.1', 
                                   gsub('\\+? years?$', '', 
                                        empLengthLevels))))
loansData$Employment.Length.Ordered = ordered(loansData$Employment.Length, 
                                              levels(reorder(empLengthLevels, empLengthNumeric)))
loansData$Employment.Length = empLengthNumeric[loansData$Employment.Length]


intRate = loansData$Interest.Rate
```
Introduction:
------------

Lending Club (https://www.lendingclub.com/) is an online site facilitating loans
from peers to peers. When someone requests a loan, Lenging Club determines the 
interest rate of the loan based on the characteristics of the person asking 
(e.g., employment history, credit history, and creditworthiness score) and of 
the loan itself (e.g., amount, purpose, and duration).

The FICO score is a popular measure of creditworthiness which incorporates data 
such as payment history, ratio of current revolving debt, and recent searches 
of credit history [^1]. As such, we anticipate a significant negative 
correlation between the interest rate and an applicant's FICO score (i.e., we
would expect the interest rate to be lower for an applicant with a higher FICO
score). Here I performed an analysis to determine if there are additional 
variables that have an important association with interest rate after taking 
into account an applicant's FICO score.

Using exploratory analysis and standard multiple regression techinques, I show
that both loan length and loan amount are associated with the interest rate of 
a loan after adjusting for the applicant's FICO score.

Methods:
-------

_Data Collection_

For our analysis, I used a sample of `r nrow(loansData)` peer-to-peer loans 
issued at Lending Club which included non-personally identifying information on
the person asking for the loan and information on the loan itself (e.g., 
amount, interest rate, length, and purpose). The data was downloaded from a
link on Coursera Data Analysis Project 1 web page [^2] on November 14, 2013 
using the R programming language. 

_Exploratory Analysis_

After examining the raw data provided, I identified transformation to perform 
such as converting the factors for FICO ranges and interest rate to numeric
values. Exploratory analysis was done by examining tables and plots of the 
data as well as comparing the effects of different data transformations in 
regression models.  This analysis was used to determine the transformations 
and terms used in the regression model relating interest rate to FICO score, 
loan length, and loan size.

_Statistical Modeling_

To relate interest rate to FICO score we performed a standard multivariate 
linear regression model. Model selection was performed on the basis of our 
exploratory analysis and prior knoweldge of the composition and range of FICO
scores. Coefficients were estimated with ordinary least squares and standard 
errors were calculated using standard asymptotic approximations.

_Reproducibility_

All Analyses performed in this manuscript are reproduced in the R markdown file
loansFinal.Rmd. 

Results:
-------

The loans data used in this analysis contains the following fields for each 
loan: its Interest Rate (IR), FICO score of the loan applicant (FICO), the 
length of the loan (Len), and the amount requested in the application (Req) 
[^3]. We identified no missing values in this dataset.

Over 50% of loans had an interest rate between 
`r paste0(collapse='-', round(quantile(intRate, c(.25,.75))))`% with just
over `r floor(100*sum(intRate > 20)/nrow(loansData))`% of loans having an 
interest rate greater than 20%.

Given that the range of FICO scores is between 300 and 850[^4], through 
exploratory analysis, I determined the best fit to interest rate to be the
inverse cube of the scaled FICO score:

$$
\begin{aligned}
 FICO.Scaled &= (FICO - 300)/(850-300) \\
 IR &= b_0 + \frac{b_1}{FICO.Scaled^3} + e_1
\end{aligned}
$$
```{r transform, include=FALSE, echo=FALSE, eval=TRUE}
ficoScaled = (loansData$FICO - 300)/(850-300)
fInv = 1/(ficoScaled^3)
```
```{r model1, include=FALSE, echo=FALSE, eval=TRUE}
model1 = lm(intRate ~ fInv)
```

where $b_0$ is an intercept term and $b_1$ represents the change in interest 
rate given a change of 1 in the inverse cube of FICO.Scaled. The error term 
$e_1$ represents the variation in interest rate unaccounted for by this 
transformation of the FICO score. 

Through additional analysis, I determined that both loan length and amount 
requested could account for some of this residual variance.  Using

$$
\begin{aligned}
IR &= b_0 + \frac{b_1}{FICO.Scaled^3} + f(Len) + {b_2}\frac{Req}{1000} + e_2
\end{aligned}
$$
```{r model2, include=FALSE, echo=FALSE, eval=TRUE}
areq = loansData$Amount.Requested/1000
model2 = lm(intRate ~ fInv + loansData$Loan.Length + areq)
s2 = summary(model2)
```
where $b_0$ is an intercept term and $b_1$ are interpreted as before; 
$f(Len)$ represent a factor model with a different level for 60 month loans
than 36 month loans;  and $b_2$ represents the change in interest rate given 
a change of \$1000 dollar in the loan amount requested. The error term $e_2$ 
represents the variation in interest rate unaccounted for by these 3 terms.

We observed a highly statistically significant 
($P \le$ `r max(s2$coefficients[,4])`) association between loan interest rate 
and each of these 3 terms.  A change of 1 unit in the inverse cube of 
FICO.Scaled corresponded to a change of $b_1=`r model2$coefficients[2]`$ in the
loan percentage (95% Confidence Interval: `r confint(model2, level=.95)[2,]`); 
A 60 month loan instead of a 36 month loan corresponded to a change of 
$f(Len)=`r model2$coefficients[3]`$ inthe loan percentage
(95% Confidence Interval: `r confint(model2, level=.95)[3,]`); and a change of 
\$1000 in the requested loan amount corresponded to a change of 
($b_2=`r model2$coefficients[4]`$) in the loan percentage 
(95% Confidence Interval: `r confint(model2, level=0.95)[4,]`). So, for 
example, if we had two applicants with the same FICO score, and the first 
requested a 36 month loan, and the second a 60 month, we would expect the 
second to have an interest rate `r model2$coefficients[3]`% higher.  If the 
second applicant requested \$1000 more, we would expect the interest rate to 
increase an additional `r model2$coefficients[4]`%

Conclusions
-----------

My analysis suggests that there are significant positive correlations between
the interest rate and both loan length and loan size even after taking into 
account FICO score. As expected, FICO score showed the strongest correlation 
to interest rate, but I was able to improve the fit by using the loan length 
and loan size.  No other features in the dataset improved the fit as much as 
these two.  One possible reason for this could be that many of the other 
features (e.g., monthly income, employment length, and revolving credit 
balance) are already factored into the FICO score.


[^1]: http://en.wikipedia.org/wiki/Credit_score_in_the_United_States

[^2]: `r dataRdaUrl`

[^3]: `r codebookUrl`

[^4]: http://en.wikipedia.org/wiki/Credit_score_in_the_United_States#FICO_score_ranges
